Conceptualizing open health data platforms for low- and middle income countries

### 2024-04-02 index.qmd
Add OpenHIM section

The main contributions of this paper is that:
- It provides a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework
- Provides details on type 1 and 2 data sharing, comparing the different design options

@@

Various studies have investigated the merits of FHIR and its performance vis-a-vis other healthcare standards. Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data]. A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR concluded that i) these three standards are functionally and technically compatible, and therefore can be used side by side; and that ii) each of these standards have their strengths and limitations that correlate with their intended use as summarized in the @tbl-comparison.

@@

We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different service layers by synthesizing the current best practices of the a lakehouse architecture of [@hai2023data;@harby2022data;@harby2024data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services). These 5 services are considered the core of the FHIR data platform, and although it practice often a downstream dashboarding or visualization component is used, that component is not considered part of this architecture as such. NB: distinction between exploration and dashboarding is not absolute.

@@

- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting
- Maintenance-related functions remain the same
- NB: orchestration falls under data provenance

@@

Strictly speaking not part of the architecture,

@@

![ Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible. The category Analytics Services is not a part of the original OpenHIE and is discussed in the paper. Point-of-Service systems are excluded for brevity. A systematic review of such digital public goods is beyond the scope of this document.](images/digital-public-goods.png){#tbl-digital-public-goods}

Allows for the Data Warehouse data to be visualised via a BI tool (Apache Superset)

@@

- One of the most used IOL layers in LMICs. Leading development of InstantHIE library to configure OpenHIE infrastructures

[https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing](https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing)

- Accept FHIR bundles submitted securely through an IOL (OpenHIM)
- Stores Clinical FHIR data to a FHIR store (HAPI FHIR)
- Stores Patient Demographic data to an MPI (JeMPI)
- Pushes FHIR resources to Kafka for the reporting pipeline (and other systems) to use
- Pulls FHIR data out of Kafka and maps it to flattened tables in the Data Warehouse (Clickhouse)

- View generation in different packages, not portable
  - Logstash for bulk
  - Kafka for streaming, uses JSON for mappings which looks similar to SQL-on-FHIR

-

@@

|    |    |
|:---|:---|
| ELT | Extract, Load and Transform |
| FAIR |Findable, Accessible, Interoperable and Reusable |
| FHIR |Fast Healthcare Interoperability Resources |
| FL | Federated learning |
| HDP	|	Health data platform, explicitly differentiated from health digital platform |
| HIE | Health Information Exchange |
| IR | Intermediate Representation |
| LMIC | Low- and middle income countries |
| PET | Privacy-enhancing technologies |

>>>>

The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; and ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture. As such, it aims to inform future developments and implementation of open digital health platforms in LMICs.

@@

Various studies have investigated the merits of FHIR and its performance vis-a-vis other healthcare standards. Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data;@tsafnat2024converge]. A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR concluded that i) these three standards are functionally and technically compatible, and therefore can be used side by side; and that ii) each of these standards have their strengths and limitations that correlate with their intended use as summarized in the @tbl-comparison.

@@

We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different service layers by synthesizing the current best practices of the a lakehouse architecture of [@hai2023data;@harby2022data;@harby2024data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services). These 5 services are considered the core of the FHIR data platform, and although in practice often a downstream dashboarding or visualization component is used, that component is not the main focus of our analysis. Rather, we aim to elucidate and conceptualize the core "FHIR Data Lakehouse".

@@

- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting !! We need to stress this, because this is an existing OpenHIE workflow
- Maintenance-related functions remain the same
- NB: orchestration falls under data provenance

@@

- Many tools, often focus on creating information dashboards and visualizations
- Compatibility with processing & API: which query languages and interfaces are supported. Some dialect of SQL, some dialect of NoSQL, dataframe API, all of the above?

@@

![Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible. The category Analytics Services is not a part of the original OpenHIE and is discussed in the paper. Point-of-Service systems are excluded for brevity. A systematic review of such digital public goods is beyond the scope of this document.](images/digital-public-goods.png){#tbl-digital-public-goods}

@@

To evaluate the extended OpenHIE architecture described above, we first consider the OpenHIM Platform. The Open Health Information Mediator (OpenHIM, [http://openhim.org/](OpenHIMhttp://openhim.org/))) component is the reference implementation of the Interoperability Layer (IOL) as defined in the OpenHIE specification. The most current version (8.4.2 at the time of writing) provides all the core functions including central point of access for the services of the HIE; routing functions; central logging for auditing and debugging purposes; and orchestration/mediation mechanisms to co-ordinate requests. This reference implementation of the IOL has recently been extended as a reference implementation of the so-called 'Instant OpenHIE' concept, that aims to address the primary needs of (i) allowing implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) having a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@digitalsquare2020instantopenhie]. The Instant HIE project allows OpenHIE components to be packages up, deployed, operated and scaled via a simple Command Line Interface (CLI, version 2 released in 2023 [https://github.com/openhie/instant-v2](https://github.com/openhie/instant-v2)). The OpenHIM Platform ([https://jembi.gitbook.io/openhim-platform](https://jembi.gitbook.io/openhim-platform)) provides a set of bundled packages (called recipes) that can be deployed using the Instant OpenHIE deployment tool. In the following we will evaluate the recipe for "a central data repository with a data warehouse" that provides "A FHIR-based Shared Health record linked to a Master Patient Index (MPI) for linking and mathing patient demographics and a default reporting pipeline to transform and visualise FHIR data" ([https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing](https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing)).

![Overview of the default data stack of the OpenHIM Platform. The default stack (top, darkblue) consists of Kafka, Clickhouse and Superset. An alternative solution based on the ELK stack is also supported (bottom, lightblue), consisting of Elasticsearch, Logstash and Kibana.](images/openhim-platform.png){#fig-openhim-platform}

@fig-openhim-platform shows a schematic overview of two data stacks that are supported in the OpenHIM platform. The Shared Health Record (SHR, implemented with HAPI FHIR server) and the Client Registry (CR, implemented with JeMPI server) are the sources that store clinical FHIR data and patient demographic data, respectively. The default data stack is based on streaming ingestion using Kafka into a Clickhouse database. As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics. Subsequently, each FHIR resource topic is flatted using mappers Unbundling and flattening of FHIR resources is done in Kafka using FHIRPath. Superset is used as the tool for consuming the data to create dashboard visualizations.

The OpenHIM platform also support data analytics based on the ELK stack, where data is ingested in bulk using Logstash, stored in Elasticsearch and made available for consumption in Kibana. Also here, the incoming FHIR bundles are unbundled in Logstash into separate FHIR resources. However, given that Elasticsearch is a document-based search engine, the FHIR resources are stored as-is with no flattening. Exploring and analysing the data requires writing queries in Elasticsearch Query Language (ES|QL), either through the query interface of Elasticsearch or using Kibana.

Evaluating these two data stacks, we see the following:

- Pattern of flattening FHIR resources with FHIRPath expressions is very close to the idea of SQL-on-FHIR. Although it doesn't adhere to this new standard in the strict sense, the philosophy of generating tabular views is the same
- When using the ELK stack, flattening is done at the end. Implementations of FHIRPath support Elasticsearch as an execution engine, also here 
- Main limitations: both Clickhouse en Elasticsearch don't follow decomposition of storage, compute and UI. Therefore, downward scaleability is limited.

@@

|    |    |
|:---|:---|
| CLI | Command-line Interface |
| CR | Client Registry |
| ELK | Elasticsearch, Logstach and Kibana stack |
| ELT | Extract, Load and Transform |
| FAIR |Findable, Accessible, Interoperable and Reusable |
| FHIR |Fast Healthcare Interoperability Resources |
| FL | Federated learning |
| HDP	|	Health data platform, explicitly differentiated from health digital platform |
| HIE | Health Information Exchange |
| IR | Intermediate Representation |
| LMIC | Low- and middle income countries |
| MPC | Multiparty Computation |
| PET | Privacy-enhancing technologies |
| SHR | Shared Health Record |

### 2024-04-02 index.tex
Add OpenHIM section

The main contributions of this paper is that: - It provides a framework
for the components of the Data \& Analysis Services that builds on
current best practices from the data engineering community into the
OpenHIE framework - Provides details on type 1 and 2 data sharing,
comparing the different design options

@@

Various studies have investigated the merits of FHIR and its performance
vis-a-vis other healthcare standards. Comparisons between OpenEHR, ISO
13606, OMOP and FHIR have been made
\citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data}.
A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR
concluded that i) these three standards are functionally and technically
compatible, and therefore can be used side by side; and that ii) each of
these standards have their strengths and limitations that correlate with
their intended use as summarized in the Table~\ref{tbl-comparison}.

@@

We propose to extend the OpenHIE architecture with a ``Data and
Analytics Services'' domain with different service layers by
synthesizing the current best practices of the a lakehouse architecture
of \citep{hai2023data, harby2022data, harby2024data} and the composable
data stack \citep{pedreira2023composable} (Figure~\ref{fig-ohie},
Table~\ref{tbl-data-and-analysis-services}). These 5 services are
considered the core of the FHIR data platform, and although it practice
often a downstream dashboarding or visualization component is used, that
component is not considered part of this architecture as such. NB:
distinction between exploration and dashboarding is not absolute.

@@

\begin{itemize}
\tightlist
\item
  SQL-on-FHIR Views provide new standard to support mADX aggregate
  reporting
\item
  Maintenance-related functions remain the same
\item
  NB: orchestration falls under data provenance
\end{itemize}

@@

Strictly speaking not part of the architecture,

@@

Allows for the Data Warehouse data to be visualised via a BI tool
(Apache Superset)

@@

\begin{itemize}
\tightlist
\item
  One of the most used IOL layers in LMICs. Leading development of
  InstantHIE library to configure OpenHIE infrastructures
\end{itemize}

\url{https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing}

\begin{itemize}
\item
  Accept FHIR bundles submitted securely through an IOL (OpenHIM)
\item
  Stores Clinical FHIR data to a FHIR store (HAPI FHIR)
\item
  Stores Patient Demographic data to an MPI (JeMPI)
\item
  Pushes FHIR resources to Kafka for the reporting pipeline (and other
  systems) to use
\item
  Pulls FHIR data out of Kafka and maps it to flattened tables in the
  Data Warehouse (Clickhouse)
\item
  View generation in different packages, not portable

  \begin{itemize}
  \tightlist
  \item
    Logstash for bulk
  \item
    Kafka for streaming, uses JSON for mappings which looks similar to
    SQL-on-FHIR
  \end{itemize}
\item
\end{itemize}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ELT & Extract, Load and Transform \\
FAIR & Findable, Accessible, Interoperable and Reusable \\
FHIR & Fast Healthcare Interoperability Resources \\
FL & Federated learning \\
HDP & Health data platform, explicitly differentiated from health
digital platform \\
HIE & Health Information Exchange \\
IR & Intermediate Representation \\
LMIC & Low- and middle income countries \\
PET & Privacy-enhancing technologies \\
\end{longtable}

>>>>

The main contributions of this paper are i) description of a framework
for the components of the Data \& Analysis Services that builds on
current best practices from the data engineering community into the
OpenHIE framework; and ii) evaluation of different implementations and
design options for type 1 and 2 data sharing within an extended OpenHIE
architecture. As such, it aims to inform future developments and
implementation of open digital health platforms in LMICs.

@@

Various studies have investigated the merits of FHIR and its performance
vis-a-vis other healthcare standards. Comparisons between OpenEHR, ISO
13606, OMOP and FHIR have been made
\citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data, tsafnat2024converge}.
A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR
concluded that i) these three standards are functionally and technically
compatible, and therefore can be used side by side; and that ii) each of
these standards have their strengths and limitations that correlate with
their intended use as summarized in the Table~\ref{tbl-comparison}.

@@

We propose to extend the OpenHIE architecture with a ``Data and
Analytics Services'' domain with different service layers by
synthesizing the current best practices of the a lakehouse architecture
of \citep{hai2023data, harby2022data, harby2024data} and the composable
data stack \citep{pedreira2023composable} (Figure~\ref{fig-ohie},
Table~\ref{tbl-data-and-analysis-services}). These 5 services are
considered the core of the FHIR data platform, and although in practice
often a downstream dashboarding or visualization component is used, that
component is not the main focus of our analysis. Rather, we aim to
elucidate and conceptualize the core ``FHIR Data Lakehouse''.

@@

\begin{itemize}
\tightlist
\item
  SQL-on-FHIR Views provide new standard to support mADX aggregate
  reporting !! We need to stress this, because this is an existing
  OpenHIE workflow
\item
  Maintenance-related functions remain the same
\item
  NB: orchestration falls under data provenance
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Many tools, often focus on creating information dashboards and
  visualizations
\item
  Compatibility with processing \& API: which query languages and
  interfaces are supported. Some dialect of SQL, some dialect of NoSQL,
  dataframe API, all of the above?
\end{itemize}

@@

@@

To evaluate the extended OpenHIE architecture described above, we first
consider the OpenHIM Platform. The Open Health Information Mediator
(OpenHIM, \href{OpenHIMhttp://openhim.org/}{http://openhim.org/}))
component is the reference implementation of the Interoperability Layer
(IOL) as defined in the OpenHIE specification. The most current version
(8.4.2 at the time of writing) provides all the core functions including
central point of access for the services of the HIE; routing functions;
central logging for auditing and debugging purposes; and
orchestration/mediation mechanisms to co-ordinate requests. This
reference implementation of the IOL has recently been extended as a
reference implementation of the so-called `Instant OpenHIE' concept,
that aims to address the primary needs of (i) allowing implementers to
engage with a preconfigured health information exchange solution and
running tools (based on the architecture) and test their applicability
and functionality with a real health context problem; and (ii) having a
packaged reference version of the OpenHIE architecture that is comprised
of a set of reference technologies and other appropriate tools that form
the building blocks of the health information exchange that can be
configured and extended to support particular use cases
\citep{digitalsquare2020instantopenhie}. The Instant HIE project allows
OpenHIE components to be packages up, deployed, operated and scaled via
a simple Command Line Interface (CLI, version 2 released in 2023
\url{https://github.com/openhie/instant-v2}). The OpenHIM Platform
(\url{https://jembi.gitbook.io/openhim-platform}) provides a set of
bundled packages (called recipes) that can be deployed using the Instant
OpenHIE deployment tool. In the following we will evaluate the recipe
for ``a central data repository with a data warehouse'' that provides
``A FHIR-based Shared Health record linked to a Master Patient Index
(MPI) for linking and mathing patient demographics and a default
reporting pipeline to transform and visualise FHIR data''
(\url{https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing}).

\includegraphics{images/openhim-platform.png}

\caption{\label{fig-openhim-platform}Overview of the default data stack
of the OpenHIM Platform. The default stack (top, darkblue) consists of
Kafka, Clickhouse and Superset. An alternative solution based on the ELK
stack is also supported (bottom, lightblue), consisting of
Elasticsearch, Logstash and Kibana.}

Figure~\ref{fig-openhim-platform} shows a schematic overview of two data
stacks that are supported in the OpenHIM platform. The Shared Health
Record (SHR, implemented with HAPI FHIR server) and the Client Registry
(CR, implemented with JeMPI server) are the sources that store clinical
FHIR data and patient demographic data, respectively. The default data
stack is based on streaming ingestion using Kafka into a Clickhouse
database. As part of the ingestion, incoming FHIR bundles that contain
multiple FHIR resources are unbundled in separate topics. Subsequently,
each FHIR resource topic is flatted using mappers Unbundling and
flattening of FHIR resources is done in Kafka using FHIRPath. Superset
is used as the tool for consuming the data to create dashboard
visualizations.

The OpenHIM platform also support data analytics based on the ELK stack,
where data is ingested in bulk using Logstash, stored in Elasticsearch
and made available for consumption in Kibana. Also here, the incoming
FHIR bundles are unbundled in Logstash into separate FHIR resources.
However, given that Elasticsearch is a document-based search engine, the
FHIR resources are stored as-is with no flattening. Exploring and
analysing the data requires writing queries in Elasticsearch Query
Language (ES\textbar QL), either through the query interface of
Elasticsearch or using Kibana.

Evaluating these two data stacks, we see the following:

\begin{itemize}
\tightlist
\item
  Pattern of flattening FHIR resources with FHIRPath expressions is very
  close to the idea of SQL-on-FHIR. Although it doesn't adhere to this
  new standard in the strict sense, the philosophy of generating tabular
  views is the same
\item
  When using the ELK stack, flattening is done at the end.
  Implementations of FHIRPath support Elasticsearch as an execution
  engine, also here
\item
  Main limitations: both Clickhouse en Elasticsearch don't follow
  decomposition of storage, compute and UI. Therefore, downward
  scaleability is limited.
\end{itemize}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CLI & Command-line Interface \\
CR & Client Registry \\
ELK & Elasticsearch, Logstach and Kibana stack \\
ELT & Extract, Load and Transform \\
FAIR & Findable, Accessible, Interoperable and Reusable \\
FHIR & Fast Healthcare Interoperability Resources \\
FL & Federated learning \\
HDP & Health data platform, explicitly differentiated from health
digital platform \\
HIE & Health Information Exchange \\
IR & Intermediate Representation \\
LMIC & Low- and middle income countries \\
MPC & Multiparty Computation \\
PET & Privacy-enhancing technologies \\
SHR & Shared Health Record \\
\end{longtable}
